Density dependence

### 2020-09-30 R/density dependence.Rmd
simplified density dependence analysis to only use 3 years of data.  Corrected calculation of density, re-ran models.

@@

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha
```

```{r}
ha <-
  ha %>% 
  group_by(ha_id_number) %>% 
  arrange(ha_id_number, year) %>% 
  ungroup() %>% 
  #scale year
  mutate(year_post = year - min(year))
```

Filter to just continuous forest

```{r}
ha_cf <- 
  ha %>% 
  filter(habitat == "CF") %>% 
  mutate(log_shts = log(shts),
         log_shts_next = log(shts_next))
```

Or filter to just 1-ha and cf

```{r}
ha_sub <-
  ha %>% 
  filter(habitat %in% c("CF", "1-ha")) %>% 
  mutate(log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat))
```

Technically, the density data are only for 2006, 2007, and 2009, so I should really only use those three transition years.  This might still not be 100% correct.

```{r}
ha_sub_sub <-
  ha_sub %>% 
  filter(year %in% c(2006, 2007, 2009))
```

@@

For each vital rate I'll include the number of shoots and habitat as fixed effects and plant ID and year as random effects.  I'm not including plot since there are few plots in each habitat type, and preliminary modeling shows the variance for plot is the lowest of those three random effects.

@@

```{r}
mpois <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
                family = poisson(link = "log"),
                data = ha_sub)

@@

mnorm <- glmer(shts_next ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
                family = gaussian(link = "log"),
                data = ha_sub)
```

@@

Lognormal distribution fits best according to AIC, although poisson has lower RMSE.  I guess despite being discrete, this isn't really "count" data *per se*.  The lognormal model doesn't have problems with singularity that the poisson model has.  Both have problems with heteroscedasticity.  I'll go with log-normal.

@@

@@

```{r}
m_growth2 <- update(m_growth, .~. - (1|ha_id_number))
AICtab(m_growth, m_growth2)
```

@@

```{r}
m_growth3 <- update(m_growth, .~. - (1|year_post))
AICtab(m_growth, m_growth2, m_growth3)
```
Can't remove year

@@

```{r}
Anova(m_growth)
fixef(m_growth)
```

Positive density dependence at 0--5 meters

Same for 3-year subset?

```{r}
mgrowth_sub <- update(m_growth, data = ha_sub_sub)
Anova(mgrowth_sub)
```

No longer significant.

@@

```{r}
m_surv <- glmer(surv ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year), 
                family = binomial, data = ha_sub)
```

@@

@@

```{r}
m_surv2 <- update(m_surv, .~. - (1|ha_id_number))
AICtab(m_surv, m_surv2)
```

@@

```{r}
m_surv3 <- glm(surv ~ log_shts + habitat + shts_0_5 + shts_5_10, 
                family = binomial, data = ha_sub)
AICtab(m_surv2, m_surv3)
```
Can't remove year.

@@

```{r}
Anova(m_surv2)
fixef(m_surv2)
```

@@

Does this hold for the 3-year subset?

```{r}
m_surv2_sub <- update(m_surv2, data = ha_sub_sub)
Anova(m_surv2_sub)
fixef(m_surv2_sub)
```

Yes, nearly identical coefs.

@@

```{r}
ggplot(ha_sub, aes(infl)) + geom_histogram()
```
Make a binary column

@@

#the way suvival is currently coded, a plant can flower and then not survive in the same year, so no need to filter out dead plants.
ha_sub %>% filter(surv == 0 & flwr == 1)
```

```{r}
m_flwr <- glmer(flwr ~ log_shts + habitat + shts_0_5 + shts_5_10 + (1|ha_id_number) + (1|year_post),
                family = binomial, data = ha_sub)
```

```{r}
m_flwr2 <- update(m_flwr, .~. - (1|ha_id_number))
AICtab(m_flwr, m_flwr2)
```
Can't remove plant ID.

```{r}
m_flwr3 <- update(m_flwr, .~. - (1|year_post))
AICtab(m_flwr, m_flwr2, m_flwr3)
```

Can't remove year

```{r}
Anova(m_flwr)
fixef(m_flwr)
```

Ok, there's an effect, but oddly only at the further distance.  Again, positive density dependence.

Probability of flowering increases with size and is lower in continuous forest

No flowering data for 2006, 2007, and 2009.  These models won't run if the data are subset to only include these years.

@@

**Not working yet**

So it seems like there are effect of density at least out to 20m.  I'd like to take the approach in Teller et al. 2016, but the code is so convoluted that it is not adaptable. I want to start by trying to adapt the `dlnm` approach instead.

@@

```{r eval=FALSE}
Q <-
  ha_sub %>%
  select(shts_0_0.5:shts_4.5_5) %>%
  as.matrix()

@@

## GAM

```{r eval=FALSE}
library(dlnm)
library(mgcv)

# Q_log <- log(Q+1)
g_growth <-
  gam(shts_next ~ s(log_shts) +
        habitat +
        s(Q, L,
          bs = "cb",
          # k = c(10, 10),
          xt = list(bs = "cr")),
      family = gaussian(link = "log"),
      data = ha_sub,
      method = "REML")
gam.check(g_growth)
anova(g_growth)
```

Marginally significant effect of conspecific density on growth

```{r eval=FALSE}
g_surv <-
  gam(surv ~ s(log_shts, k = 12) +
        habitat +
        s(Q, L,
          bs = "cb",
          # k = c(10, 10),
          xt = list(bs = "cr")),
      family = binomial,
      data = ha_sub,
      method = "REML")
gam.check(g_surv)
anova(g_surv)
```
No significant effect of density

## Predict

```{r eval=FALSE}
hist(Q)
pred_growth <- crosspred("Q", model = g_growth, cumul = TRUE, cen = mean(Q, na.rm = T), from = 0, to = 8)
summary(pred_growth)
```

## Plot crossbasis

### Contour plot

```{r}
plot.crosspred(pred_growth,
               ptype = "contour",
               xlab = "#shoots/m^2",
               ylab = "distance (m)",
               cumul = FALSE,
               main = "Relative to mean competition")
```
Non-monotonic density dependence that decays steeply beyond 1m

### Slices

We can also plot slices through the contour plot at different distances or different shoot densities

```{r eval=FALSE}
plot.crosspred(pred_growth, ptype = "slice", var = 0.2, xlab = "distance (m)", main = "0.2 shoots/m^2", ylim = c(0.9,1.1))
plot.crosspred(pred_growth, ptype = "slice", var = 1, xlab = "distance (m)", main = "1 shoots/m^2", ylim = c(0.9,1.1))
plot.crosspred(pred_growth, ptype = "slice", var = 2, xlab = "distance (m)", main = "2 shoots/m^2", ylim = c(0.9,1.1))
plot.crosspred(pred_growth, ptype = "slice", var = 8, xlab = "distance (m)", main = "8 shoots/m^2", ylim = c(0.9,1.1))
```

```{r}
plot.crosspred(pred_growth, ptype = "slice", lag = 0.25, xlab = "density (shoots/m^2)", main = "0.25m", ylim = c(0.9,1.1))
plot.crosspred(pred_growth, ptype = "slice", lag = 1.25, xlab = "density (shoots/m^2)", main = "1.25m", ylim = c(0.9,1.1))
plot.crosspred(pred_growth, ptype = "slice", lag = 3.25, xlab = "density (shoots/m^2)", main = "3.25m", ylim = c(0.9,1.1))
plot.crosspred(pred_growth, ptype = "slice", lag = 4.25, xlab = "density (shoots/m^2)", main = "4.25m", ylim = c(0.9,1.1))
```

## Other shapes

It is odd that the strength of competition seems to be increasing with distance.  I'd like to be able to force the distance dimension to be monotonically decreasing as was done in Teller et al (I think?).  Teller does this *after* fitting a spline using `mk_dist_wts()`, which is a function that takes the spline coefficients, and predicted values from the spline, and transforms it into a non-negative, monotonic decreasing function.

You can override one of the dimensions of the cross basis with any custom function.

```{r eval=FALSE}
#use arglag to override lag dimension of crossbasis, use argvar to override plant density dimension.
lin <- dlnm:::lin
xt <- list(bs = "cr", arglag = list(fun = "lin"))

g_growth <- 
  gam(shts_next ~ s(log_shts) +
        habitat +
        s(Q, L,
          bs = "cb",
          k = c(5, 4),
          fx = c(F, T), #turns off penalization for the lag dimension
          xt = xt),
      family = gaussian(link = "log"),
      data = ha_sub,
      gamma = 1.5)

gam.check(g_growth)
```

```{r eval=FALSE}
x <- crosspred("Q", model = g_growth)
plot.crosspred(x, ptype = "contour", xlab = "#shoots/100m^2", ylab = "distance (m)")

plot.crosspred(x, ptype = "slice", var = 100)
plot.crosspred(x, ptype = "slice", var = 10)
```
Now the kernel is the same shape at every value of density.  However, this relationship doesn't make any sense.  Fewer knots?  More penalty?  Custom function?

```{r eval=FALSE}
anova(g_growth)
```

# TODO:

- ~~is density at 0 distance including the focal plant?  It shouldn't.~~
- Check number of inflorescences given flowered?
- ~~log-transform density (nah, not right now.  DOesn't help with convergence)~~
- ~~check that survival is coded right so that it is a function of `ht_prev`~~
- try negative binomial distribution for shoot number
- ~~Density should be total shoot number~~
- ~~Should actually be density in previous year~~
- ~~Shoots should be poisson distribution, not log-transformed?~~
- ~~incorporate 1-ha fragments and use habitat type as a predictor.~~
- Implement FLM approach to density dependence.

>>>>

# TODO

- Get confidence intervals for cross-basis predictions.  Plot slices with confidence bands.

@@

```{r data, echo=TRUE}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha
```
This is the main *Heliconia* dataset, but with densities of conspecifics (shoots/m^2^) at different rings with inner and outer radii in meters.  So, for example, the `shts_8_8.5` column is the number of *Heliconia* shoots/m^2^ in a ring 8 to 8.5 m around the focal plant.

The `x` and `y` coordinates are averages of data from 2006, 2007, and 2009, but joined to the full range of the dataset--assuming that plants don't move.

## Tidying and subsetting

```{r}
ha <-
  ha %>%
  ungroup() %>% 
  arrange(ha_id_number, year) %>% 
  #scale year, and convert to factor since I'll be using as random effect
  mutate(year_post = as_factor(year - min(year)))
```

Use just 1-ha and cf

```{r}
ha_sub <-
  ha %>% 
  filter(habitat %in% c("CF", "1-ha")) %>% 
  mutate(log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat),
         ha_id_number = as.factor(ha_id_number))
```

@@

For each vital rate I'll include the number of shoots and habitat as fixed effects and plant ID and year as random effects. For the mixed effects models I'll use density at just two rings (0--5m and 5--10m) as fixed effects. There are probably too few plots (4 1-ha and 6CF) to use plot as a random effect, but I don't know how to code it as a fixed effect right now.

```
vital_rate ~ log(shts) + habitat + shts_0_2.5 + shts_2.5_5 + (1|year) + (1|ha_id_number) + (1|habitat:plot)
```

@@

```{r}
mpois <-
  glmer(
    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
    family = poisson(link = "log"),
    data = ha_sub
  )

@@

mnorm <-
  glmer(
    shts_next ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
    family = gaussian(link = "log"),
    data = ha_sub
  )
```

@@

Lognormal distribution fits best according to AIC, although poisson has lower RMSE.  I guess despite being discrete, this isn't really "count" data *per se*.  The lognormal model doesn't have problems with singularity that the poisson model has.  Both have problems with heteroscedasticity. A log-transformation of the response doesn't fix the heteroscedatsticyt either. I'll go with log-normal.

@@

Plot:

```{r}
m_growth2 <- update(m_growth, .~. - (1|habitat:plot))
AICtab(m_growth, m_growth2)
```

@@

```{r}
m_growth3 <- update(m_growth2, .~. - (1|ha_id_number))
AICtab(m_growth2, m_growth3)
```

@@

```{r}
m_growth4 <- update(m_growth2, .~. - (1|year_post))
AICtab(m_growth, m_growth2, m_growth3, m_growth4)
```
Can remove year

@@

```{r}
Anova(m_growth4)
fixef(m_growth4)
```

No density dependence

@@

```{r}
m_surv <-
  glmer(
    surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot), 
    family = binomial, data = ha_sub
  )
```

@@

Plot:

```{r}
m_surv2 <- update(m_surv, .~. - (1|habitat:plot))
AICtab(m_surv, m_surv2)
```

@@

```{r}
m_surv3 <- update(m_surv2, .~. - (1|ha_id_number))
AICtab(m_surv2, m_surv3)
```

@@

```{r}
m_surv4 <- glm(surv ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5, 
                family = binomial, data = ha_sub)
AICtab(m_surv4, m_surv3)
```
glm is best

@@

```{r}
Anova(m_surv4)
coef(m_surv4)
```

@@

@@

Make a binary column

@@

unique(ha_sub$flwr)
```

Can't do this model because no variation in flowering.

```{r eval=FALSE}
m_flwr <-
  glmer(
    flwr ~ log_shts + habitat + shts_0_2.5 + shts_2.5_5 +
      (1|ha_id_number) + (1|year_post) + (1|habitat:plot),
    family = binomial, data = ha_sub
  )
```

@@

I'd like to take the approach in Teller et al. 2016, but the code is so convoluted that it is not adaptable. Instead, I'll start by trying to adapt the `dlnm` approach.

@@

```{r}
Q <-
  ha_sub %>%
  select(shts_0_0.5:shts_4.5_5) %>%
  as.matrix()

@@

```{r}
library(dlnm)
library(mgcv)
library(visibly) # better diagnostic visulazations.  one of several packages for this.
source(here("R", "utils.R"))
```

## Growth

```{r}
g_growth <-
  gam(shts_next ~ 
        s(log_shts) + #Essentially linear, could replace with parametric
        habitat +
        s(Q, L,
          bs = "cb",
          xt = list(bs = "cr")), #+
        # s(ha_id_number, bs = "re"), #this takes waaay too long to converge
      family = gaussian(link = "log"),
      data = ha_sub,
      method = "REML")
```

```{r}
plot_gam_check(g_growth)
```

```{r}
anova(g_growth)
```

Effect of size on size(t+1):

```{r}
plot(g_growth, pages = 1)
```

Effects of density:

```{r}
pred_growth <- pred_cb(Q, L, g_growth)
ggplot(pred_growth, aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("#shoots (t+1)", option = "A")
```

## Survival

```{r}
g_surv <-
  gam(surv ~
        s(log_shts) +
        habitat +
        s(Q, L,
          bs = "cb",
          # k = c(10, 10),
          xt = list(bs = "cr")),
      family = binomial,
      data = ha_sub,
      method = "REML")
```

```{r}
# plot_gam_check(g_surv)
anova(g_surv)
```

Effect of size on survival:

```{r}
plot(g_surv, pages = 1)
```

```{r}
pred_surv <- pred_cb(Q, L, g_surv)
ggplot(pred_surv, aes(x = x, y = lag, z = fitted, fill = fitted)) +
  geom_raster(interpolate = TRUE) +
  geom_contour(color = "grey50", size = 0.4) + 
  scale_y_continuous("distance (m)", expand = c(0,0)) +
  scale_x_continuous("conspecific density (shoots/m^2)", expand = c(0,0)) +
  scale_fill_viridis_c("P(survival)", option = "A")
```
